{\large Master Thesis}\\{\normalsize on

### 2011-07-20 doc/report.tex
m

@@

One way of solving the LCE problem without any preprocessing uses $O(|\LCE(i,j)|)$ query time. We call this algorithm \proc{DirectComp}. For a query $\LCE(i,j)$, the algorithm compares $s[i]$ with $s[j]$, then $s[i+1]$ with $s[j+1]$ and so on, until the two characters differ, or the end of the string is reached. The worst case query time is $O(n)$ on a string of length $n$, and the average case query time is $O(1)$, because the average case LCE value over all possible inputs of a given string of length $n$ and alphabet size $\sigma$ is $O(1/(\sigma-1))=O(1)$~\cite{ilie-navarro-tinta}.

@@

The other method, which we call \proc{LcpRmq}, uses constant time Range Minimum Queries on a Longest Common Prefix array. The LCP array contains the length of the longest common prefixes of each pair of neighbor suffixes in SA. The length of the longest common prefix of two arbitrary suffixes in SA can be found as the minimum of all LCP values of neighbor suffixes between the two desired suffixes, because SA lists the suffixes in lexicographical ordering, i.e. $\LCE(i,j)=\LCP[\RMQ_{\LCP}(\SAinv[i] + 1, \SAinv[j])]$, where $\SAinv[i] < \SAinv[j]$. % An example is shown in \pref{fig:sa+lcp+min}.

@@

\begin{corollary}
\fprint[2] has $O(n)$ space and $O(\sqrt n)$ query time. It has two levels, where one uses a table of fingerprints and the other uses the original string.
\end{corollary}

\begin{corollary}
\fprint[\logceil] has $O(n\cdot\log n)$ space and $O(\log n)$ query time. The data structure is equivalent to the one generated by Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg}, and a query only needs to do one comparison at each level.
\end{corollary}

@@

Existing state of the art solutions are either good in worst case, while poor in average case (\proc{LcpRmq}), or good in average case while poor in worst case (\proc{DirectComp}). Our \fprintk\ solution targets a worst case vs. average case query time tradeoff between these two extremes. \pref{fig:wc-avg} illustrates the relative query time performances of these solutions. Our solution is almost as fast as \proc{DirectComp} on an average case input, and it is a lot faster than \proc{DirectComp} on a worst case input. Compared to \proc{LcpRmq}, our solution has a significantly better performance on an average case input, but its worst case performance is not as good as that of \proc{LcpRmq}. The space usage for \proc{LcpRmq} and \fprintk\ are approximately the same when $k=6$.

@@

\paragraph{Tree notation.}
The ancestors of a node $u$ is $u$ itself as well as any node, which is a parent of an ancestor of $u$. The depth of a node $u$ is one greater than the depth of $u$'s parent, and the depth of the root is zero. A path in a tree begins at node $u$ and ends at node $v$, where $u$ is an ancestor of $v$.

\paragraph{Sorting complexity.} The time it takes to sort $n$ numbers within an universe of $\sigma$ is written as $\sortt$.

@@

There are $n$ leaves in a suffix tree, and since each internal node has at least two children, the tree has less than $2n$ nodes. Each edge uses $O(1)$ space, as the label is a substring of $s$ and it can therefore be represented by two indexes of $s$. The suffix tree therefore takes $O(n)$ space.

@@

\paragraph{SA and LCP.} For a string $s$ of length $n$ with alphabet $\sigma$, the Suffix Array, \SA, is an array of length $n$, which encodes the lexicographical ordering of all suffixes of $s$. The lexicographically smallest suffix is $\suff_{\SA[1]}$, the lexicographically largest suffix is $\suff_{\SA[n]}$, and the lexicographically $i$'th smallest suffix is $\suff_{\SA[i]}$. The inverse Suffix Array, \SAi, describes where a given suffix is in the lexicographical order. Suffix $\suff_i$ is the lexicographically $\SAinv[i]$'th smallest suffix.

@@

\paragraph{Predecessor.}
Given a set of $n$ objects, where the key of each object is a natural number between one and $U$, the predecessor of a natural number $x$ is an object with maximal key less than or equal to $x$. We can solve the static Predecessor problem in $O(\log\log U)$ time using $O(n)$ space~\cite{predecessor}.

@@

\paragraph{Heavy Path Decomposition}
The weight of a node $u$ is the size of the subtree rooted at $u$. For a given node $u$, a child of greatest weight is marked heavy, and all other children are marked light. The root is always marked light. An edge going from a node to a heavy child node is heavy, and an edge going from a node to a light child node is light. A path of only heavy edges is a heavy path, and the combined length of all heavy paths in a tree of $n$ nodes is less than $n$. The markings of heavy and light edges and nodes in a tree is a Heavy Path Decomposition. Every simple path in a tree of size $n$ contains at most $O(\log n)$ light edges, and we can construct a Heavy Path Decomposition in $O(n)$ time~\cite{nca}.

\paragraph{Ladder Decomposition}
To make a ladder decomposition of a tree, we first create a ladder of nodes from the root of the tree to the leaf of greatest depth. We then remove the nodes in the ladder from the tree and repeat this operation recursively for all remaining subtrees, until all nodes are part of a ladder. Node $u$'s ladder is the ladder which contains $u$ at this point. Finally, we double the number of nodes in each ladder by extending it towards the root of the tree. If we reach the root before we have doubled the number of nodes, we stop at the root. Each node may thus be in multiple ladders.

@@

LCE of two indexes $i$ and $j$ is defined as the length of the longest common prefix of $\suff_i$ and $\suff_j$. In a suffix tree, the path from the root to $L_i$ has label $\suff_i$ (likewise for $j$). The path label from the root to the nearest common ancestor of $L_i$ and $L_j$ is therefore a common prefix of the two suffixes. Since two child edges of the same node never have the same first character on their edge labels, the path label from the root to $\NCA_\mathcal{T}(L_i,L_j)$ is not only a common prefix, but longest common prefix of $\suff_i$ and $\suff_j$. We can therefore find the LCE value as $\LCE_s(i,j) = D[\NCA_\mathcal{T}(L_i,L_j)]$, where $\mathcal{T}$ is the suffix tree for $s$, and $D[v]$ is the label depth of node $v$ in the tree.

@@

The LCP array contains the length of the longest common prefixes of each pair of neighbor suffixes in SA. The length of the longest common prefix of two arbitrary suffixes in SA can be found as the minimum of all LCP values of neighbor suffixes between the two desired suffixes, because SA lists the suffixes in lexicographical ordering, i.e. $\LCE(i,j)=\LCP[\RMQ_{\LCP}(\SAinv[i] + 1, \SAinv[j])]$, where $\SAinv[i] < \SAinv[j]$. An example is shown in \pref{fig:sa+lcp+min}. Preprocessing time is $O(\sortt)$, space requirement is $O(n)$ and query time is $O(1)$.

@@

Different algorithms exists for solving the RMQ problem, and we can solve it in optimal $O(1)$ time and $O(n)$ space. But like for the LCE problem, the asymptotically best RMQ solution may not be the best RMQ solution in practice. We examine the following RMQ solutions, where \RMQpq{p}{q} has $O(p)$ space and preprocessing time, and $O(q)$ query time:
\begin{samepage}
\begin{description}
\item[\RMQn] walks through the array directly in $O(n)$ time and does not use any preprocessing (except for transforming LCE into RMQ).
\item[\RMQq] preprocesses the array in $O(n)$ space in order to do queries in $O(1)$ time. This is a twolevel data structure as shown in \pref{fig:sa+lcp+min-twolevel}. The array is split into blocks of size $O(\log n)$. The top level is an array of length $O(n/\log n)$, where each element contains the minimum value one block. The top level uses a \RMQpq{n\log n}{1}-solution, and the lower level precomputes all answers for each block, which is done in $O(n)$ space, since there are only a limited number of different block types of size $O(\log n)$.
\item[\RMQlog] is the same as \RMQq\ except that the lower level is replaced with a direct search through a $O(\log n)$ length block.
\end{description}
\end{samepage}

@@

\begin{lemma}
The worst case query time for \fprintk\ is $O(k\cdot n^{1/k})$, and the average case query time is $O(1)$.
\end{lemma}
\begin{proof}
Step 1 takes $O(k)$ time. In step 2 and 3, the number of remaining characters left to check at level $\ell$ is $O(n^{(\ell+1)/k})$, since the previous level found two differing substrings of that length (at the top level $\ell=k-1$ we have $O(n^{(\ell+1)/k}) = O(n)$). Since we can check $t_\ell = \Theta(n^{\ell/k})$ characters in constant time at level $\ell$, the algorithm uses $O(n^{(\ell+1)/k})/\Theta(n^{\ell/k}) = O(n^{1/k})$ time at that level. Over all $k$ levels, $O(k\cdot n^{1/k})$ query time is used.

At each step except step 3, the algorithm increments $v$. Step 3 is executed the same number of times as step 1, in which $v$ is incremented. The query time is therefore linear to the number of times $v$ is incremented, and it is thereby $O(v)$. The query time is thus $O(1)$ in average case, where $v=O(1)$.
\end{proof}

@@

We could have left out step 1 of the query algorithm in \pref{sec:fingerprint-query} and started with $\ell=k-1$, which would change the query pattern as shown in \pref{fig:fingerprint-updown}. This would keep the asymptotic worst case query time of $O(k\cdot n^{1/k})$, but it would increase our average case query time to $O(k)$. Our experiments have shown that whenever $k$ is small, keeping step 1 improves average case query time, while it does not have a measurable effect on worst case query times. When $k = \logceil$, keeping step 1 can double the worst case query time, while it can make the average case query time 40 times faster for an input string of ten million characters. We want to optimize our LCE query time for the average case where the LCE value is small, so our results in \pref{sec:results} does not include the variant of \fprintk\ with $O(k)$ average case query time.

In step 1 of our query algorithm we perform one comparison at each level. We could instead do up to $O(n^{1/k})$ comparisons at each level without affecting asymptotic times. Our experiments have shown that always doing one comparison at each level is the best in practice.

@@

@@

The amount if I/O's used by \fprintk\ is $O(k\cdot n^{1/k})$. However if we structure our tables of fingerprints differently, we can improve the number of I/O's to $O(k(n^{1/k}/B+1))$ in the cache-oblivious model. Instead of storing the fingerprint of $s[i\twodots i+t_\ell]$ at $H_\ell[i]$, we can store it at $H_\ell[((i-1)\mod t_\ell)\cdot\lceil n/t_\ell\rceil+\lfloor (i-1)/t_\ell\rfloor+1]$. This will group all used fingerprints at level $\ell$ next to each other in memory, such that the amount of I/O's at each level is reduced from $O(n^{1/k})$ to $O(n^{1/k}/B)$.

@@

We implemented different variants of \proc{DirectComp}, \proc{LcpRmq}, \proc{DirectCompLookup} and \fprintk\ in C++ and compared them to each other. We have not implemented \proc{SuffixNca}, since others have shown that \proc{LcpRmq} is better than \proc{SuffixNca} in practice~\cite{ilie-navarro-tinta}. The algorithms we compared are the following:
\begin{description}
\item[\proc{DirectComp}] is the simple \proc{DirectComp} algorithm with no preprocessing and worst case $O(n)$ query time.
\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac] is our new \fprintk\ algorithm using $k$ levels, where $k$ is $2$, $3$ and $\logceil$. The numbers \textless$t_{k-1}$, ..., $t_1$\textgreater\ describe the exact size of fingerprinted substrings at each level.
\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater wc] is a variant of our new \fprintk\ algorithm, which starts at level $\ell=k-1$ instead of $\ell=0$, as discussed in \pref{sec:improved-avg}.
\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac\textless$u_0$, ..., $u_{k-2}$\textgreater] is a variant of our new \fprintk\ algorithm, which stays at level $\ell$ until $v\geq u_\ell$, as discussed in \pref{sec:improved-avg}, instead of doing one comparison at each level in step 1 of the query in \pref{sec:fingerprint-query}.
\item[\proc{RMQ}\textless$n$, $1$\textgreater] is the \proc{LcpRmq} algorithm using constant time RMQ.
\item[\proc{RMQ}\textless$p$, $q$\textgreater$_{\textrm{virtual}}$] is the \proc{LcpRmq} algorithm using different RMQ algorithms. They all contain an additional virtual method call to make implementation simpler, but their results cannot be directly compared to other algorithms, which do not have this extra virtual method call.
\item[\proc{DirectComp}\textless$c$\textgreater\proc{RMQ}\textless$n$, $1$\textgreater] is the \proc{DirectComp+LcpRmq} algorithm, where the cutoff from \proc{DirectComp} to \proc{LcpRmq} is after $c$ iterations.
\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac$_{\textrm{cache-horizontal-}\textit{op}}$] is a cache optimized variant of our \fprintk\ algorithm, where \textit{op} describes which of the shift or multiplication arithmetic operations it uses.
\end{description}

@@

For each kind of strings, we tested the algorithms using the following pattern:
\begin{enumerate}
\item Generate a string of length $n$ and generate a million random pairs $(i, j)$, where each of $i$ and $j$ is chosen between $1$ and $n$ uniformly at random.
\item For each tested algorithm, preprocess the given string, and run a query for each of the million pairs. Measure the time it takes to run all the queries combined.
\item Double the value of $n$ and repeat from step 1.
\end{enumerate}

@@

\pref{fig:plot-rmq-directcomp} also shows \proc{DirectComp+LcpRmq}, where we first try \proc{DirectComp}, and if we have not found the LCE value after one iteration, we use \proc{LcpRmq} instead. We see that \proc{DirectComp+LcpRmq} is worse than \proc{DirectComp} on average case strings. We could increase the cutoff from one iteration to something bigger, which might improve average case performance. That would however increase worst case time. None of the two plots we measure represents a good worst case for \proc{DirectComp+LcpRmq} compared to the other algorithms we test. The worst case for \proc{DirectComp+LcpRmq} relative to other the other algorithms would be when the average LCE value is slightly larger than the cutoff, such that \proc{DirectComp+LcpRmq} reaches \proc{LcpRmq} in most queries, while other algorithms still work on LCE values as small as possible. As described in \pref{sec:dc-lcprmq}, \proc{DirectComp+LcpRmq} also uses a lot of space.

@@

In \pref{fig:plot-fingerprint-avglog} we see that on worst case strings, the query time of \fprint[\logceil] degrade a small amount when using the average case optimization. Both variations of \fprint[\logceil] show query times, which are $O(\log n)$ as expected. On average case strings, the query time of \fprint[\logceil] improves significantly when using average case optimization, and both variants show their expected $O(1)$ and $O(\log n)$ average case query times.

@@

In \pref{fig:plot-fingerprint-avg3l} we see that doing one comparison at each level on the way up is better than doing $n^{1/k}$ comparisons. This is most visible on worst case strings, where the algorithm doing one comparison at each level uses less time to reach level $\ell=k-1$, which it will reach anyway. On average case strings, the code specialized in handling only one comparison at each level is simpler. On the question of doing average case optimization or not, we see the same results for \fprint[3], as we saw for \fprint[\logceil], but at a smaller scale.

@@

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-cache-rand10.plt}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-cache-alla.plt}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-cache-repeat-pow.plt}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-cache-rand2.plt}
    \end{center}
    \caption{\label{fig:plot-fingerprint-cache-horiz}Query times of \fprint[2] with and without cache optimization.}
\end{figure}

@@

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-log-cache-rand10.plt}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-log-cache-alla.plt}
    \end{center}
    \caption{\label{fig:plot-fingerprint-cache-vert}Query times of \fprint[2] with and without cache optimization.}
\end{figure}

@@

To test practical performance, we have implemented the algorithms in C++. The implementation is based on that of Ilie~et~al.~\cite{ilie-navarro-tinta}, which we have heavily modified. We have made it compile on a 32-bit Windows architecture, and generalized it to allow more algorithms and tests to be implemented. \pref{tab:test-variables} summarizes the structure of the code and the parameters we can vary.

\begin{table}[tp]
\centering
\begin{tabular}{l l l}
Variable & Example & Component \\ \hline
alphabet size & $\sigma = 10$ & string \\
character pattern & uniformly at random & string \\ \hline
algorithm & \fprintk & algorithm \\
alg. parameter & $t = 2\sqrt n$ & algorithm \\ \hline
string length & exponentially increasing from $1$ to $100.000$ & test \\
query pattern & uniformly at random & test \\ \hline
\end{tabular}
\caption{List of performance variables tested.}\label{tab:test-variables}
\end{table}

>>>>

% TODOs
% skarpere plots
% hav plot for $r=\sqrt n$ med til forsvar

@@

One way of solving the LCE problem without any preprocessing uses $O(|\LCE(i,j)|)$ query time. We call this algorithm \proc{DirectComp}. For a query $\LCE(i,j)$, the algorithm compares $s[i]$ to $s[j]$, then $s[i+1]$ to $s[j+1]$ and so on, until the two characters differ, or the end of the string is reached. The worst case query time is $O(n)$ on a string of length $n$, and the average case query time is $O(1)$, because the average case LCE value over all possible inputs of a given string of length $n$ and alphabet size $\sigma$ is $O(1/(\sigma-1))=O(1)$~\cite{ilie-navarro-tinta}.

@@

The other method, which we call \proc{LcpRmq}, uses constant time Range Minimum Queries on a Longest Common Prefix array. The LCP array contains the length of the longest common prefixes of each pair of neighbor suffixes in SA. The length of the longest common prefix of two arbitrary suffixes in SA can be found as the minimum of all LCP values of neighbor suffixes between the two desired suffixes, because SA lists the suffixes in lexicographical ordering. I.e. $\LCE(i,j)=\LCP[\RMQ_{\LCP}(\SAinv[i] + 1, \SAinv[j])]$, where $\SAinv[i] < \SAinv[j]$.

@@

\begin{corollary}
\fprint[2] uses $O(n)$ space and $O(\sqrt n)$ query time. It has two levels, where one uses a table of fingerprints and the other uses the original string.
\end{corollary}

\begin{corollary}
\fprint[\logceil] uses $O(n\cdot\log n)$ space and $O(\log n)$ query time. The data structure is equivalent to the one generated by Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg}, and a query only needs to do one comparison at each level.
\end{corollary}

@@

In practice, existing state of the art solutions are either good in worst case, while poor in average case (\proc{LcpRmq}), or good in average case while poor in worst case (\proc{DirectComp}). Our \fprintk\ solution targets a worst case vs. average case query time tradeoff between these two extremes. \pref{fig:wc-avg} illustrates the relative query time performances of these solutions. Our solution is almost as fast as \proc{DirectComp} on an average case input, and it is significantly faster than \proc{DirectComp} on a worst case input. Compared to \proc{LcpRmq}, our solution has a significantly better performance on an average case input, but its worst case performance is not as good as that of \proc{LcpRmq}. The space usage for \proc{LcpRmq} and \fprintk\ are approximately the same when $k=6$.

@@

\paragraph{Tree notation.}
The ancestors of a node $u$ is $u$ itself as well as any node, which is a parent of an ancestor of $u$. A path in a tree begins at node $u$ and ends at node $v$, where $u$ is an ancestor of $v$. The length of a path is the number of edges between nodes in the path. The depth of a node $u$ is the length of the path from the root to $u$, and the height of $u$ is the length of the longest path from $u$ to a descendant of $u$. E.g. the depth of the root and the height of a leaf is both zero.

\paragraph{Sorting complexity.} The time it takes to sort $n$ numbers within an universe of size $\sigma$ is written as $\sortt$.

@@

There are $n$ leaves in a suffix tree, and since each internal node has at least two children, the tree has less than $2n$ nodes. Each edge uses $O(1)$ space, since the label is a substring of $s$ and it can therefore be represented by two indexes of $s$. The suffix tree therefore takes $O(n)$ space.

@@

\paragraph{SA and LCP.} For a string $s$ of length $n$ with alphabet size $\sigma$, the Suffix Array, \SA, is an array of length $n$, which encodes the lexicographical ordering of all suffixes of $s$. The lexicographically smallest suffix is $\suff_{\SA[1]}$, the lexicographically largest suffix is $\suff_{\SA[n]}$, and the lexicographically $i$'th smallest suffix is $\suff_{\SA[i]}$. The inverse Suffix Array, \SAi, describes where a given suffix is in the lexicographical order. Suffix $\suff_i$ is the lexicographically $\SAinv[i]$'th smallest suffix.

@@

\paragraph{Predecessor.}
Given a set of $n$ objects, where the key of each object is a natural number between one and $U$, the predecessor of a natural number $x$ is an object with maximal key, whose key is at most $x$. We can solve the static Predecessor problem in $O(\log\log U)$ time using $O(n)$ space~\cite{predecessor}.

@@

\paragraph{Heavy Path Decomposition}
The weight of a node $u$ is the size of the subtree rooted at $u$. For each node $u$, a child of greatest weight is marked heavy, and all other children are marked light. The root is always marked light. An edge going from a node to a heavy child node is heavy, and an edge going from a node to a light child node is light. A path of only heavy edges is a heavy path, and the combined length of all heavy paths in a tree of $n$ nodes is less than $n$. The markings of heavy and light edges and nodes in a tree is a Heavy Path Decomposition. Every simple path in a tree of size $n$ contains at most $O(\log n)$ light edges, and we can construct a Heavy Path Decomposition in $O(n)$ time~\cite{nca}.

\paragraph{Ladder Decomposition}
To make a ladder decomposition of a tree, we first create a ladder of nodes on the path from the root of the tree to the leaf of greatest depth. We then remove the nodes in the ladder from the tree and repeat this operation recursively for all remaining subtrees, until all nodes are part of a ladder. Node $u$'s ladder is the ladder which contains $u$ at this point. Finally, we double the number of nodes in each ladder by extending it towards the root of the tree. If we reach the root before we have doubled the number of nodes, we stop at the root. Each node may thus be in multiple ladders.

@@

LCE of two indexes $i$ and $j$ is defined as the length of the longest common prefix of $\suff_i$ and $\suff_j$. In a suffix tree, the path from the root to $L_i$ has label $\suff_i$ (likewise for $j$). The path label on the path from the root to the nearest common ancestor of $L_i$ and $L_j$ is therefore a common prefix of the two suffixes. Since two child edges of the same node never have the same first character on their edge labels, the path label on the path from the root to $\NCA_\mathcal{T}(L_i,L_j)$ is not only a common prefix, but the longest common prefix of $\suff_i$ and $\suff_j$. We can therefore find the LCE value as $\LCE_s(i,j) = D[\NCA_\mathcal{T}(L_i,L_j)]$, where $\mathcal{T}$ is the suffix tree for $s$, and $D[v]$ is the label depth of node $v$ in the tree.

@@

The LCP array contains the length of the longest common prefixes of each pair of neighbor suffixes in SA. The length of the longest common prefix of two arbitrary suffixes in SA can be found as the minimum of all LCP values of neighbor suffixes between the two desired suffixes, because SA lists the suffixes in lexicographical ordering. I.e. $\LCE(i,j)=\LCP[\RMQ_{\LCP}(\SAinv[i] + 1, \SAinv[j])]$, where $\SAinv[i] < \SAinv[j]$. An example is shown in \pref{fig:sa+lcp+min}. Preprocessing time is $O(\sortt)$, space requirement is $O(n)$ and query time is $O(1)$.

@@

Different algorithms exists for solving the RMQ problem, and we can solve it in optimal $O(1)$ time and $O(n)$ space. But like for the LCE problem, the asymptotically best RMQ solution may not be the best RMQ solution in practice. We examine the following RMQ solutions, where \RMQpq{p}{q} has $O(p)$ space and preprocessing time, and $O(q)$ query time:
\begin{samepage}
\begin{description}
\item[\RMQn] walks through the array directly in $O(n)$ time and does not use any preprocessing (except for transforming LCE into RMQ).
\item[\RMQq] preprocesses the array in $O(n)$ space in order to do queries in $O(1)$ time. This is a twolevel data structure as shown in \pref{fig:sa+lcp+min-twolevel}. The array is split into blocks of size $O(\log n)$. The top level is an array of length $O(n/\log n)$, where each element contains the minimum value of one block. The top level uses a \RMQpq{n\log n}{1}-solution, and the lower level precomputes all answers for each block, which is done in $O(n)$ space, since there are only a limited number of different block types of size $O(\log n)$.
\item[\RMQlog] is the same as \RMQq\ except that the lower level is replaced with a direct search through a $O(\log n)$ length block.
\end{description}
\end{samepage}

@@

\begin{lemma}
The worst case query time for \fprintk\ is $O(k\cdot n^{1/k})$, and the average case query time is $O(1)$.
\end{lemma}
\begin{proof}
Step one takes $O(k)$ time. In step two and three, the number of remaining characters left to check at level $\ell$ is $O(n^{(\ell+1)/k})$, since the previous level found two differing substrings of that length (at the top level $\ell=k-1$ we have $O(n^{(\ell+1)/k}) = O(n)$). Since we can check $t_\ell = \Theta(n^{\ell/k})$ characters in constant time at level $\ell$, the algorithm uses $O(n^{(\ell+1)/k})/\Theta(n^{\ell/k}) = O(n^{1/k})$ time at that level. Over all $k$ levels, $O(k\cdot n^{1/k})$ query time is used.

At each step except step three, the algorithm increments $v$. Step three is executed the same number of times as step one, in which $v$ is incremented. The query time is therefore linear to the number of times $v$ is incremented, and it is thereby $O(v)$. The query time is thus $O(1)$ in average case, where $v=O(1)$.
\end{proof}

@@

We could have left out step one of the query algorithm in \pref{sec:fingerprint-query} and started with $\ell=k-1$, which would change the query pattern as shown in \pref{fig:fingerprint-updown}. This would keep the asymptotic worst case query time of $O(k\cdot n^{1/k})$, while it might improve practical worst case query time, but it would increase our average case query time to $O(k)$.
\ifarticle
Our experiments have shown that whenever $k$ is small, keeping step one improves average case query time, while it does not have a measurable effect on worst case query times. When $k = \logceil$, keeping step one can double the worst case query time, while it can make the average case query time 40 times faster for an input string of ten million characters. We want to optimize our LCE query time for the average case where the LCE value is small, so our results in \pref{sec:results} does not include the variant of \fprintk\ with $O(k)$ average case query time.
\fi % article

@@

In step one of our query algorithm we perform one comparison at each level. We could instead do up to $O(n^{1/k})$ comparisons at each level without affecting asymptotic times.
\ifarticle
Our experiments have shown that always doing one comparison at each level is the best in practice.
\fi % article

@@

The amount if I/O's used by \fprintk\ is $O(k\cdot n^{1/k})$. However if we structure our tables of fingerprints differently, we can improve the number of I/O's to $O(k(n^{1/k}/B+1))$ in the cache-oblivious model. Instead of storing the fingerprint of $s[i\twodots i+t_\ell-1]$ at $H_\ell[i]$, we can store it at $H_\ell[((i-1)\mod t_\ell)\cdot\lceil n/t_\ell\rceil+\lfloor (i-1)/t_\ell\rfloor+1]$. This will group all used fingerprints at level $\ell$ next to each other in memory, such that the amount of I/O's at each level is reduced from $O(n^{1/k})$ to $O(n^{1/k}/B)$.

@@

We implemented different variants of \proc{DirectComp}, \proc{LcpRmq}, \proc{DirectCompLookup} and \fprintk\ in C++ and compared them to each other. We have not implemented \proc{SuffixNca}, since others have shown that \proc{LcpRmq} is better than \proc{SuffixNca} in practice~\cite{ilie-navarro-tinta}. The algorithms we compared are the following:
\begin{description}
\item[\proc{DirectComp}] is the simple \proc{DirectComp} algorithm with no preprocessing and worst case $O(n)$ query time.
\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac] is our new \fprintk\ algorithm using $k$ levels, where $k$ is $2$, $3$ and $\logceil$. The numbers \textless$t_{k-1}$, ..., $t_1$\textgreater\ describe the exact size of fingerprinted substrings at each level.
\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater wc] is a variant of our new \fprintk\ algorithm, which starts at level $\ell=k-1$ instead of $\ell=0$, as discussed in \pref{sec:improved-avg}.
\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac\textless$u_0$, ..., $u_{k-2}$\textgreater] is a variant of our new \fprintk\ algorithm, which stays at level $\ell$ until $v\geq u_\ell$, as discussed in \pref{sec:improved-avg}, instead of doing one comparison at each level in step one of the query in \pref{sec:fingerprint-query}.
\item[\proc{RMQ}\textless$n$, $1$\textgreater] is the \proc{LcpRmq} algorithm using constant time RMQ.
\item[\proc{RMQ}\textless$p$, $q$\textgreater$_{\textrm{virtual}}$] is the \proc{LcpRmq} algorithm using different RMQ algorithms. They all contain an additional virtual method call to make implementation simpler, but their results cannot be directly compared to other algorithms, which do not have this extra virtual method call.
\item[\proc{DirectComp}\textless$c$\textgreater\proc{RMQ}\textless$n$, $1$\textgreater] is the \proc{DirectComp+LcpRmq} algorithm, where the cutoff from \proc{DirectComp} to \proc{LcpRmq} is after $c$ iterations.
\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac$_{\textrm{cache}-\textit{dir}-\textit{op}}$] is a cache optimized variant of our \fprintk\ algorithm, where \textit{dir} describes which of our two cache optimized variants is used, and \textit{op} describes which of the shift or multiplication arithmetic operations it uses.
\end{description}

@@

For each kind of strings, we tested the algorithms using the following pattern:
\begin{enumerate}
\item Generate a string of length $n$ and generate a million random pairs $(i, j)$, where each of $i$ and $j$ is chosen between $1$ and $n$ uniformly at random.
\item For each tested algorithm, preprocess the given string, and run a query for each of the million pairs. Measure the time it takes to run all the queries combined.
\item Double the value of $n$ and repeat from step one.
\end{enumerate}

@@

\pref{fig:plot-rmq-directcomp} also shows \proc{DirectComp+LcpRmq}, where we first try \proc{DirectComp}, and if we have not found the LCE value after one iteration, we use \proc{LcpRmq} instead. We see that \proc{DirectComp+LcpRmq} is worse than \proc{DirectComp} on average case strings. We could increase the cutoff from one iteration to something bigger, which might improve average case performance. That would however increase worst case time. None of the two plots we measure represents a good worst case for \proc{DirectComp+LcpRmq} compared to the other algorithms we test. The worst case for \proc{DirectComp+LcpRmq} relative to the other algorithms would be when the average LCE value is slightly larger than the cutoff, such that \proc{DirectComp+LcpRmq} reaches \proc{LcpRmq} in most queries, while other algorithms still work on LCE values as small as possible. As described in \pref{sec:dc-lcprmq}, \proc{DirectComp+LcpRmq} also uses a lot of space.

@@

In \pref{fig:plot-fingerprint-avglog} we see that on worst case strings, the query time of \fprint[\logceil] doubles when we use the average case optimization. On average case strings, the query time of \fprint[\logceil] is up to 40 times faster when using average case optimization. All four cases in \pref{fig:plot-fingerprint-avglog} show their expected $O(1)$ and $O(\log n)$ query times. In \pref{fig:plot-fingerprint-avg3l}, we see the same results for \fprint[3], as we saw for \fprint[\logceil], but at a smaller scale. The negative effect of doing average case optimization is barely measurable on worst case strings, while the improvement on average case strings is up to 40\%.

@@

In \pref{fig:plot-fingerprint-avg3l} we see that doing one comparison at each level on the way up in \fprint[3] is better than doing $n^{1/k}$ comparisons. This is most visible on worst case strings, where the algorithm doing one comparison at each level uses less time to reach level $\ell=k-1$, which it will reach anyway. On average case strings, the code specialized in handling only one comparison at each level is simpler.

@@

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-cache-rand10.plt}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-cache-alla.plt}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-cache-repeat-pow.plt}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-cache-rand2.plt}
    \end{center}
    \caption{\label{fig:plot-fingerprint-cache-horiz}Query times of \fprint[2] with and without intra-level cache optimization.}
\end{figure}

@@

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-log-cache-rand10.plt}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-log-cache-alla.plt}
    \end{center}
    \caption{\label{fig:plot-fingerprint-cache-vert}Query times of \fprint[2] with and without inter-level cache optimization.}
\end{figure}

@@

To test practical performance, we have implemented the algorithms in C++. The implementation is based on that of Ilie~et~al.~\cite{ilie-navarro-tinta}, which we have heavily modified. We have added a working build system, made it compile on a 32-bit Windows architecture, and generalized it to allow more algorithms and tests to be implemented. \pref{tab:test-variables} summarizes the structure of the code and the parameters we can vary.

\begin{table}[tp]
\centering
\begin{tabular}{l l l}
Variable & Example & Component \\ \hline
alphabet size & $\sigma = 10$ & string \\
character pattern & uniformly at random & string \\ \hline
algorithm & \fprintk & algorithm \\
alg. parameters & $t_1 = 2\sqrt n$ & algorithm \\ \hline
string length & exponentially increasing from $1$ to $100.000$ & test \\
query pattern & uniformly at random & test \\ \hline
\end{tabular}
\caption{List of performance variables tested.}\label{tab:test-variables}
\end{table}

In our implementation, we have made the following assumptions about the LCE problem:
\begin{itemize}
\item A character in a string is always eight bits.
\item A string always ends with a special null character, which does not occur anywhere else in the string. This does not affect our results, since we do not measure practical preprocessing time. If an algorithm has faster query times when the string ends in a special character, it could just add that character to the end of the string in its preprocessing.
\item We make no assumptions on the order of $i$ and $j$. In some applications of the LCE problem, for example approximate string searching~\cite{approx-search}, we know that $i<j$, and some algorithms might be able to take advantage of this knowledge to get faster query times. We do not measure this case.
\item Whenever we use random characters or query pairs, we use the \texttt{rand()} function from C. This gives repeatable pseudo-random numbers, which makes it easier for us to compare test runs.
\end{itemize}
